---
title: "Introduction to pixieweb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pixieweb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`pixieweb` is an R package for *discovering*, *inspecting* and *downloading* statistical data from [PX-Web](https://www.scb.se/px-web) APIs — the platform used by Statistics Sweden (SCB), Statistics Norway (SSB), Statistics Finland, and many other national statistics agencies. This vignette provides an overview of the methods included in the `pixieweb` package and the design principles of the package API. To learn more about the specifics of functions and to see a full list of the functions included, please see the [Reference section of the package homepage](https://lchansson.github.io/pixieweb/reference/index.html) or run `??pixieweb`. For a quick introduction to the package see the vignette [Quick start guide to pixieweb](a-quickstart.html).

> **Note on pxweb:** The excellent [pxweb](https://cran.r-project.org/package=pxweb)
> package by rOpenGov already provides comprehensive R access to
> PX-Web APIs. `pixieweb` is not a replacement — it offers an
> *alternative paradigm* built around search-then-fetch discovery and
> progressive disclosure. Choose the workflow that fits your needs.

The design of `pixieweb` functions is inspired by the design and functionality provided by several packages in the `tidyverse` family. pixieweb uses the base R pipe (`|>`) throughout. Some vignette examples use `dplyr`, `tidyr`, and `ggplot2` for data wrangling and visualisation:

```{r install, eval = FALSE}
install.packages("pixieweb")
```

## PX-Web, a platform for national statistical databases

[PX-Web](https://www.scb.se/px-web) is the statistical database platform used by national statistics agencies across the Nordic countries and beyond. Each agency runs its own instance — Statistics Sweden at [scb.se](https://www.scb.se), Statistics Norway at [ssb.no](https://www.ssb.no), Statistics Finland at [stat.fi](https://www.stat.fi), and many more — but they all share the same underlying API, which comes in two versions:

- **v1** — the legacy API, still widely deployed. POST-only data queries, no search endpoint, table discovery requires walking a folder hierarchy.
- **v2** — the modern API, launched by SCB in 2024. GET+POST data queries, full-text search, codelists endpoint, server-side saved queries.

`pixieweb` handles both versions transparently — the user-facing functions have the same signatures, and only the internal request building differs.

To get started with PX-Web you might want to visit the web frontend of any participating agency, or read through the [PxWebApi2 documentation](https://github.com/PxTools/PxApiSpecs) (English). However, you can also use the `pixieweb` package to explore data without prior knowledge of the database.

```{r setup}
library("pixieweb")
```

### The data model

Data in PX-Web are stored as *multi-dimensional data cubes*. Each agency publishes hundreds or thousands of **tables**, and each table defines its own set of **variables** (dimensions) along which observations are indexed. A typical population table might, for instance, be indexed by region, sex, age and year — while a foreign trade table might be indexed by partner country, commodity group and month.

When downloading data, the user needs to specify which values to include along each variable — or omit the variable entirely, in which case the API returns a pre-computed aggregate (this is called *elimination*). Every table has its own rules for which variables are *mandatory* and which are *eliminable*.

In summary, a PX-Web table is organised along four basic concepts:

- An **API instance** — a particular agency's PX-Web database (e.g. SCB, SSB)
- A **table** — a single statistical data cube published by that agency
- A set of **variables** — the dimensions of the cube (region, time, sex, product, etc.); each variable has a set of valid **values**
- One or more **content codes** — what is actually being measured in each cell (population count, deaths, tax revenue, GDP, ...)

Tables with multiple content codes can return data in either *long* format (one row per observation, one column for the content code) or *wide* format (one column per content code) — see the section on wide output below.

Additionally, many variables come with one or more **codelists** — alternative groupings of the values that can be used to aggregate data on the fly. For example, a "Region" variable with 290 municipalities might offer codelists that group them into 21 counties, or into eight NUTS-2 regions.

`pixieweb` provides a function family for each of these concepts:

| Level        | Discover               | Search             | Describe             | Extract               | Values                |
|--------------|------------------------|--------------------|----------------------|-----------------------|-----------------------|
| API instance | `px_api_catalogue()`   | —                  | —                    | —                     | —                     |
| Table        | `get_tables()`         | `table_search()`   | `table_describe()`   | `table_extract_ids()` | —                     |
| Variable     | `get_variables()`      | `variable_search()`| `variable_describe()`| `variable_extract_ids()`| `variable_values()` |
| Codelist     | `get_codelists()`      | —                  | `codelist_describe()`| `codelist_extract_ids()`| `codelist_values()` |
| Data         | `get_data()`           | —                  | —                    | —                     | —                     |

### Connecting to an API

All `pixieweb` workflows start by connecting to a PX-Web instance. `px_api()` accepts a short alias (`"scb"`, `"ssb"`, `"statfi"`, ...) or a full URL:

```{r api_mock, eval = FALSE}
# Known aliases
scb <- px_api("scb", lang = "en")
ssb <- px_api("ssb", lang = "en")

# Or a custom URL
custom <- px_api("https://my.statbank.example/api/v2/", lang = "en")

# See all known APIs
px_api_catalogue()
```

```{r api, echo = FALSE}
scb <- pixieweb:::vd_scb
catalogue <- pixieweb:::vd_catalogue
catalogue
```

For cross-country comparison and a fuller tour of the multi-API catalogue, see `vignette("multi-api")`.

### Discovering tables: `get_tables()`

Tables are the central entity. On v2 APIs, `get_tables()` sends a server-side search query; on v1 APIs, it walks the folder tree. In both cases the result is a tibble with rich metadata, one row per table:

```{r tables_mock, eval = FALSE}
# Search the SCB catalogue for population-related tables
tables <- get_tables(scb, query = "population")

head(tables)
```

```{r tables, echo = FALSE}
tables <- pixieweb:::vd_tables
tibble::tibble(head(tables))
```

The table tibble includes subject path, time period range, time unit, and data source — all of which are searchable by `table_search()` (a client-side filter, analogous to `kpi_search()` in `rKolada`).

```{r table_search}
# Narrow down to tables about municipalities
tables |>
  table_search("municipal") |>
  table_describe(max_n = 2, format = "md", heading_level = 4)
```

`table_describe()` prints a human-readable summary of each table. As with all `describe` functions in the `pixieweb`/`rKolada`/`rTrafa` family, you can set `format = "md"` to produce Markdown output that can be embedded directly in an R Markdown document by setting the chunk option `results = 'asis'`.

### Exploring variables: `get_variables()`

Once you have chosen a table, inspect its variables with `get_variables()`:

```{r variables_mock, eval = FALSE}
vars <- get_variables(scb, "TAB638")
vars |> variable_describe()
```

```{r variables, echo = FALSE}
vars <- pixieweb:::vd_variables
vars |> variable_describe()
```

Each variable has a number of important properties worth knowing about:

- **`elimination`** — can this variable be left out of your `get_data()` call? If `TRUE`, omitting the variable means the API returns a pre-computed total (e.g. omitting "Sex" gives the total for all sexes). If `FALSE`, the variable is **mandatory** and must be included in the query.
- **`time`** — is this the time dimension of the cube? (There is always exactly one.)
- **`values`** — the set of available codes and their human-readable labels.
- **`codelists`** — alternative groupings of the values (see the "Codelists" section below).

To inspect the available values for a specific variable, use `variable_values()`:

```{r variable_values_mock, eval = FALSE}
vars |> variable_values("Region")
```

```{r variable_values, echo = FALSE}
pixieweb:::vd_region_values
```

### Downloading data: `get_data()`

`get_data()` is the workhorse function. Provide an API object, a table ID, and one argument per variable you want to include in the query:

```{r get_data_mock, eval = FALSE}
pop <- get_data(scb, "TAB638",
  Region = c("0180", "1480", "1280"),
  ContentsCode = "*",
  Tid = px_top(5)
)

head(pop)
```

```{r get_data, echo = FALSE}
pop <- pixieweb:::vd_pop
tibble::tibble(head(pop))
```

Variables you **omit** are *eliminated* (aggregated) if the API allows it. If a variable is mandatory, you must include it — `get_data()` will raise an informative error otherwise.

#### Selection helpers

Often you don't want to hardcode specific codes. `pixieweb` provides a family of selection helpers that translate common wishes into API-compatible arguments:

| Helper            | Meaning                          | API versions |
|-------------------|----------------------------------|--------------|
| `c("0180")`       | Specific values                  | v1, v2       |
| `"*"`             | All values                       | v1, v2       |
| `px_top(5)`       | First N values (e.g. most recent)| v1, v2       |
| `px_bottom(3)`    | Last N values                    | v2 only      |
| `px_from("2020")` | From value onward                | v2 only      |
| `px_to("2023")`   | Up to value                      | v2 only      |
| `px_range(a, b)`  | Inclusive range                  | v2 only      |

### Metadata-driven queries: `prepare_query()`

An alternative approach to downloading data using known codes is to use the metadata tables to construct queries. For interactive exploration, `prepare_query()` inspects the table's variable metadata and builds a query with sensible defaults, so that you only need to specify the variables you care about:

```{r prepare_query_mock, eval = FALSE}
q <- prepare_query(scb, "TAB638",
  Region = c("0180", "1480", "1280")
)
q
```

```{r prepare_query, echo = FALSE}
q <- pixieweb:::vd_prepared_query
q
```

The default strategy is:

- **Content code**: all values (`"*"`)
- **Time variable**: latest 10 periods (`px_top(10)`)
- **Eliminable variables**: omitted (API aggregates)
- **Small mandatory variables** (≤ 22 values): all (`"*"`)
- **Large mandatory variables**: first value (`px_top(1)`)

With `maximize_selection = TRUE`, the function expands unspecified variables to include as many values as possible while staying under the API's cell limit. Once you're happy, pass the validated query to `get_data()`:

```{r get_data_from_query, eval = FALSE}
pop <- get_data(scb, query = q)
```

## Codelists

Codelists provide alternative groupings of variable values. They are useful when you want data at a different aggregation level than the table's default. For example, a "Region" variable with 290 Swedish municipalities might have a codelist that groups them into 21 counties (`vs_RegionLän07`):

```{r codelists_mock, eval = FALSE}
cls <- get_codelists(scb, "TAB638", "Region")
cls |> codelist_describe(max_n = 3)
```

```{r codelists, echo = FALSE}
cls <- pixieweb:::vd_codelists
cls |> codelist_describe(max_n = 3)
```

Apply a codelist in a query via the `.codelist` argument:

```{r codelist_use, eval = FALSE}
# Fetch data aggregated to counties instead of municipalities
get_data(scb, "TAB638",
  Region = "*",
  Tid = px_top(5),
  ContentsCode = "*",
  .codelist = list(Region = "vs_RegionLän07")
)
```

## Wide output and multiple content codes

When a table has multiple content codes (e.g. both Population and Deaths), the default long format has one row per content code per observation. Use `.output = "wide"` to pivot the content codes into separate columns — useful when you want to compute with several measures (e.g. death rate = Deaths / Population):

```{r wide_mock, eval = FALSE}
demo <- get_data(scb, "TAB638",
  Region = "0180",
  Tid = px_top(5),
  ContentsCode = "*",
  .output = "wide"
)
demo
```

```{r wide, echo = FALSE}
pixieweb:::vd_wide
```

## Visualising results

To explore results we can plot the downloaded data using `ggplot2`. Note that we convert the time column to a `Date` first — years as integer can produce awkward ggplot breaks like "2020, 2022.5, 2025", and a proper `Date` column lets `scale_x_date()` place tick marks on whole years. This is a pattern you will want to use for any time-series analysis across all three sibling packages (`rKolada`, `pixieweb`, `rTrafa`):

```{r plot, fig.width = 7, fig.height = 4}
library("ggplot2")

pop_plot <- pop |>
  # Keep only the Population content code (the table also has
  # "Population growth"); convert year to Date for nice axis breaks
  dplyr::filter(ContentsCode == "BE0101N1") |>
  dplyr::mutate(year = as.Date(paste0(Tid, "-01-01")))

ggplot(pop_plot, aes(year, value, colour = Region_text)) +
  # One line per region; linetype adds distinction in B/W print
  geom_line(aes(linetype = Region_text), linewidth = 1) +
  geom_point(size = 2) +
  # Years as dates, one tick per year
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  # Colour-blind-friendly palette
  scale_colour_viridis_d(option = "B", end = 0.8) +
  labs(
    title = "Population, Sweden's three most populous municipalities",
    x = "Year",
    y = "Population",
    colour = NULL,
    linetype = NULL,
    caption = px_cite(pop)
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

> **More on ggplot2?** See <https://ggplot2-book.org/>.

## Advanced: query composition

For full control over the HTTP request — useful for debugging or when you need to inspect/modify the exact query before sending it — use the low-level query composers:

```{r compose, eval = FALSE}
q <- compose_data_query(scb, "TAB638",
  Region = c("0180"),
  ContentsCode = "*",
  Tid = px_top(3)
)

# Inspect the query
q$url
q$body

# Modify and execute
raw <- execute_query(scb, q$url, q$body)
```

## Saved queries (v2 only)

PX-Web v2 supports server-side stored queries — useful for recurring reports. Save a query once, then retrieve it by ID later:

```{r saved, eval = FALSE}
# Save a query
id <- save_query(scb, "TAB638",
  Region = "0180",
  Tid = px_top(5),
  ContentsCode = "*"
)

# Retrieve later
get_saved_query(scb, id)
```

## Citation

Always cite your data sources. `px_cite()` generates a citation string from the metadata attached to a downloaded data frame:

```{r cite_mock, eval = FALSE}
px_cite(pop)
```

```{r cite, echo = FALSE}
pixieweb:::vd_citation
```

## Related packages

`pixieweb` is part of a family of R packages for Swedish and Nordic
open statistics that share the same design philosophy:

- [rKolada](https://lchansson.github.io/rKolada/) — R client for the
  [Kolada](https://kolada.se/) database of Swedish municipal and
  regional Key Performance Indicators
- [rTrafa](https://lchansson.github.io/rTrafa/) — R client for the
  [Trafa](https://api.trafa.se/) API of Swedish transport statistics

See also [pxweb](https://cran.r-project.org/package=pxweb) — the
original and established PX-Web client for R, by rOpenGov.
